---
title: "p8105_hw2_sm4956"
author: Shivangi Deepak Mewada
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 0
* created repository and R project for HW2, created rmd file and rending to GitHub. 
* created a sub-directory/ data folder that has all the data set files to be used for this HW

```{r load_libraries}
library(tidyverse)
library(readxl)
options(tibble.print_min = 5)
```

## Problem 1

reading and cleaning data, retaining important variables

```{r}
trans_ent = 
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) %>% 
  janitor::clean_names() %>% 
  select(
    line, station_name, station_latitude, station_longitude, 
    starts_with("route"), entry, exit_only, vending, entrance_type, 
    ada) %>% 
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))
```

To check the number of distinct stations:

```{r}
trans_ent %>% 
  select(station_name, line) %>% 
  distinct
```

To check the number of distinct stations that are ADA Compliant:

```{r}
trans_ent %>% 
  filter(ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

To see the proportion of station entrances/exits without vending allow entrance:

```{r}
trans_ent %>% 
  filter(vending == "NO") %>% 
  pull(entry) %>% 
  mean
```

To check the distinct stations serving A trains (and the ones that are ADA Compliant):

```{r}
trans_ent %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A") %>% 
  select(station_name, line) %>% 
  distinct

trans_ent %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A", ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

### Answers

* There are 465 distinct stations.
* There are 84 distinct stations, that are ADA compliant.
* the proportion of station entrances / exits without vending allow entrance is 0.38.
* There are 60 distinct stations, that serve the A train.
* There are 17 distinct stations, that serve the A train and are ADA compliant.

